Convergence rate of MAP estimates \\ for the exponential family

### 2021-01-28 main.tex
Expand related work

\paragraph{TODO}
The question remains of : when is such a bound valid? For which values of $\sigma^2$. To answer such questions, we need to turn to examples.

The other article \citet{lu2019relative} derives another convergence rate for SMD under another gradient variance assumption.

\section{Variance bound for centered gaussians}
Sections on proximal point and mirror descent gave us the variance assumption 
\begin{align}
	\Cov(T(x_{n}) , \natp_{n+1}) = \Cov_X \bigg(T(X), \nabla \conj (\lr_n T(X) + (1-\lr_n)\mu_n)\bigg	) 
	\leq \lr_n M \; .
\end{align}
on the sufficient statistic and $\conj$. Unfortunately this bound is not trivially satisfied, even for simple exponential family members, such as centered gaussians $\cN(0,\sigma^2)$, when we are aiming to estimate the variance.
The density of a centered normal variable is
\begin{align}
	X \sim \cN(0,\sigma^2) \implies p(x) = \inv{\sqrt{2\pi \sigma^2}} e^{-\frac{x^2}{2 \sigma^2}} \; .
\end{align}
Using the $X^2$ as a sufficient statistic we get the variance as a mean parameter and minus half the precision as a natural parameter
\begin{align}
	\quad T(X)= X^2 \ ; \quad  \mu = \sigma^2 \ ; \quad   \natp = -\inv{2 \sigma^2} = -\inv{2 \mu}\; .
\end{align}
Now we can match the log-likelihood with the exponential family template to get the log-partition function.
\begin{align}
	-\log p(x) = \frac{x^2}{2 \sigma^2} - \half \log(2\inv{2\sigma^2}) + \half \log(2 \pi) 
	= -x^2 \natp + \logpart(\natp) \\
	\implies \logpart (\natp) = -\half \log(-\natp)  + \half \log(\pi)
\end{align}
We can then use the formula $\conj(\mu) = \mu \natp - \logpart(\natp)$ to get the entropy
\begin{align}
	\conj(\mu) = -\half \log(\mu) + \half \log\frac{\pi}{2} - \half \\
\end{align}
Remark that both the entropy and the log-partition have the same shape $z\mapsto -\inv{2z}$. Which yields the same shape of Bregman divergence, as visible below (all three lines are equal)
\begin{align}
	\KL( \sigma_n^2 ; \sigma_*^2) 
	&= \half \left ( \frac{\sigma_*^2}{ \sigma_n^2} - 1 - \log \frac{\sigma_*^2}{ \sigma_n^2} \right) \\
	\bregmanconj( \mu_*; \mu_n) 
	&= \half \left ( \frac{\mu_*}{ \mu_n} - 1 - \log  \frac{\mu_*}{ \mu_n} \right) \\
	\bregman( \natp_n; \natp_* ) 
	&=  \half \left ( \frac{ \natp_n}{\natp_*} - 1 - \log  \frac{ \natp_n}{\natp_*} \right) \; .
\end{align}
In other words, this is divergence measures the discrepancy between the ratio $\frac{ \natp_n}{\natp_*} =  \frac{\mu_*}{ \mu_n}  $ and $1$ via the function $\phi(z)= \half (z - 1 - \log(z))$.
\begin{align}
	\phi(z) := \half (z - 1 - \log(z)) 
	\quad \text{ and } \quad
	\frac{ \natp_n}{\natp_*} =  \frac{\mu_*}{ \mu_n}  \\
	\implies \bregman( \natp_n; \natp_* )   = \phi(\frac{ \natp_n}{\natp_*}) =  \phi(\frac{\mu_*}{ \mu_n})
\end{align}
The covariance  \eqref{eq:hypvariance} we want to bound is equal to
\begin{align}
	\half \expect{\frac{\sigma_*^2 - X^2 }{ X^2 \lr + (1-\lr) \sigma_0^2 }}
	\approx \lr (1-\lr)^{-2} \left (\frac{\sigma_*}{\sigma_0} \right)^4
\end{align}
using a first order approximation of the inverse, eg a first order approximation of $\nabla \conj$ that holds when $\lr X^2 \ll (1-\lr) \sigma_0^2$. This does not hold for $\sigma_0\ll 1$, which is precisely when the covariance  explodes, as illustrated in Figure \TODO{}.

@@

\section{Related Work}
Exponential families became mainstream in the machine learning community to generalize frameworks such as linear regression \citep{mccullagh1989generalized} or PCA \citep{collins2001generalization} to diverse data types and distributions.
\citet[Chapter 3]{wainwright2008graphical} give a great overview of exponential families and their dual structure. 
\citet{agarwal2010geometric} highlight that the MAP problem of exponential families is also a Bregman median problem. They then use this equivalence to justify the use of conjugate prior in Bayesian estimation and hybrid generative-discriminative modeling.
Note that this equivalence holds only if sufficient statistics are within $\dom \conj  = \mathrm{Im} (\nabla\logpart)$, a point ignored by authors. 
\citet{raskutti2015information} highlight the equivalence between mirror descent in $\natp$ and natural gradient descent in $\meanp$ (or vice-versa) for the exponential family. Then they use the optimality result by Shunichi Amari on Natural Gradient Descent to show that Mirror Descent iterates reach the Cramer-Rao lower bound.
\citet{kunstner2020homeomorphic} show with an exquisite elegance that Expectation-Maximization in exponential families can be cast as mirror descent on the negative log-likelihood. Then they use recent mirror descent rates based on the idea of relative smoothness \citep{birnbaum2011distributed, bauschke2017descent, lu2018relatively} to show convergence of EM in KL-divergence, including for the ubiquitous Gaussian Mixture model, which had resisted bounding attempts for 50 years.

>>>>

\paragraph{Open questions}
The question remains of : when is such a bound valid?  what does $\sigma^2$ depend on ? To answer these questions and more, we need to turn to examples, see  Section~\ref{sec:gaussian-variance}.

\TODO{This bound might not hold in general, but it can along the trajectory. Can we prove it using an inductive method such as \citet[Appendix A]{lacoste2012simpler}?
\citet{lu2019relative} derive another convergence rate for SMD under another relative continuity and gradient variance assumption. Can we exploit it ?}

\section{Variance bound for centered gaussians}
\label{sec:gaussian-variance}
Sections on proximal point and mirror descent gave us the variance assumption 
\begin{align}
	\Cov(T(x_{n}) , \natp_{n+1}) = \Cov_X \bigg(T(X), \nabla \conj (\lr_n T(X) + (1-\lr_n)\mu_n)\bigg	) 
	\leq \lr_n M \; .
\end{align}
on the sufficient statistic and $\conj$. Unfortunately this bound is not trivially satisfied, even for simple exponential family members, such as centered gaussians $\cN(0,\sigma^2)$, when we are aiming to estimate the variance.
The density of a centered normal variable is
\begin{align}
	X \sim \cN(0,\sigma^2) \implies p(x) = \inv{\sqrt{2\pi \sigma^2}} e^{-\frac{x^2}{2 \sigma^2}} \; .
\end{align}
Using the $X^2$ as a sufficient statistic we get the variance as a mean parameter and minus half the precision as a natural parameter
\begin{align}
	\quad T(X)= X^2 \ ; \quad  \mu = \sigma^2 \ ; \quad   \natp = -\inv{2 \sigma^2} = -\inv{2 \mu}\; .
\end{align}
Now we can match the log-likelihood with the exponential family template to get the log-partition function.
\begin{align}
	-\log p(x) = \frac{x^2}{2 \sigma^2} - \half \log(2\inv{2\sigma^2}) + \half \log(2 \pi) 
	= -x^2 \natp + \logpart(\natp) \\
	\implies \logpart (\natp) = -\half \log(-\natp)  + \half \log(\pi)
\end{align}
We can then use the formula $\conj(\mu) = \mu \natp - \logpart(\natp)$ to get the entropy
\begin{align}
	\conj(\mu) = -\half \log(\mu) + \half \log\frac{\pi}{2} - \half \\
\end{align}
Remark that both the entropy and the log-partition have the same shape $z\mapsto -\inv{2z}$. Which yields the same shape of Bregman divergence, as visible below (all three lines are equal)
\begin{align}
	\KL( \sigma_n^2 ; \sigma_*^2) 
	&= \half \left ( \frac{\sigma_*^2}{ \sigma_n^2} - 1 - \log \frac{\sigma_*^2}{ \sigma_n^2} \right) \\
	\bregmanconj( \mu_*; \mu_n) 
	&= \half \left ( \frac{\mu_*}{ \mu_n} - 1 - \log  \frac{\mu_*}{ \mu_n} \right) \\
	\bregman( \natp_n; \natp_* ) 
	&=  \half \left ( \frac{ \natp_n}{\natp_*} - 1 - \log  \frac{ \natp_n}{\natp_*} \right) \; .
\end{align}
In other words, this is divergence measures the discrepancy between the ratio $\frac{ \natp_n}{\natp_*} =  \frac{\mu_*}{ \mu_n}  $ and $1$ via the function $\phi(z)= \half (z - 1 - \log(z))$.
\begin{align}
	\phi(z) := \half (z - 1 - \log(z)) 
	\quad \text{ and } \quad
	\frac{ \natp_n}{\natp_*} =  \frac{\mu_*}{ \mu_n}  \\
	\implies \bregman( \natp_n; \natp_* )   = \phi(\frac{ \natp_n}{\natp_*}) =  \phi(\frac{\mu_*}{ \mu_n})
\end{align}
The covariance  \eqref{eq:hypvariance} we want to bound is equal to
\begin{align}
	\half \expect{\frac{\sigma_*^2 - X^2 }{ X^2 \lr + (1-\lr) \sigma_0^2 }}
	\approx \lr (1-\lr)^{-2} \left (\frac{\sigma_*}{\sigma_0} \right)^4
\end{align}
using a first order approximation of the inverse, eg a first order approximation of $\nabla \conj$ that holds when $\lr X^2 \ll (1-\lr) \sigma_0^2$. This does not hold for $\sigma_0\ll 1$, which is precisely when the covariance  explodes, as illustrated in Figure \TODO{}.

@@

\section{Related Work}
Exponential families are a mainstream tool in machine learning. They are used  to generalize linear regression \citep{mccullagh1989generalized}, PCA \citep{collins2001generalization} or k-means \citep{banerjee2005clustering} to diverse data types and distributions.
\citet[Chapter 3]{wainwright2008graphical} give a great overview of exponential families and their dual structure. 
\citet{agarwal2010geometric} highlight that the MAP problem of exponential families is also a Bregman median problem. They then use this equivalence to justify the use of conjugate prior in Bayesian estimation and hybrid generative-discriminative modeling.
Note that this equivalence holds only if sufficient statistics are within $\dom \conj  = \mathrm{Im} (\nabla\logpart)$, a point ignored by authors yet critical as it does not hold for multivariate normals with unknown covariance. 
\citet{raskutti2015information} highlight the equivalence between mirror descent in $\natp$ and natural gradient descent in $\meanp$ (or vice-versa) for the exponential family. Then they use the optimality result by Shunichi Amari on Natural Gradient Descent to show that Mirror Descent iterates reach the Cramer-Rao lower bound.
\citet{kunstner2020homeomorphic} show with an exquisite elegance that Expectation-Maximization in exponential families can be cast as mirror descent on the negative log-likelihood. Then they use recent mirror descent rates based on the idea of relative smoothness \citep{birnbaum2011distributed, bauschke2017descent, lu2018relatively} to show convergence of EM in KL-divergence, including for the ubiquitous Gaussian Mixture model, which had resisted bounding attempts for 50 years.
